---
title: 'Homework 1 report'
author: "20170361 Yourim Shin"
output: 
  pdf_document:
    latex_engine: xelatex
---
## Import data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read}
clinical<-readRDS("clinical.rds") 
expression<-readRDS("expression.rds") 
mutation<-readRDS("mutation.rds")

library(dplyr)
```

## Preprocessing and normalization
```{r preprocessing}
sample_id_filter<-intersect(colnames(expression), 
                            intersect(unique(mutation$sample_id),
                                      clinical$sample_id[which(!is.na(clinical$subtype))])) 
clinical_filter<-dplyr::filter(clinical,sample_id %in% sample_id_filter)
mutation_filter<-dplyr::filter(mutation, sample_id %in% sample_id_filter) 
expression_filter<-expression[,sample_id_filter]

normalize<-function(x){
  return((x-min(x))/(max(x)-min(x)))
}

library(preprocessCore) 

tmp<-normalize.quantiles(expression_filter)
rownames(tmp)<-rownames(expression_filter)
colnames(tmp)<-colnames(expression_filter)

expression_filter<-apply(expression_filter, 1, normalize) 
expression_filter<-t(expression_filter) 
expression_filter<-na.omit(expression_filter)

#quantile_norm<-normalize.quantiles(expression_filter)
```

## Problem 1
### 1-1
```{r p1-1}
# ANOVA : aov(survival_time(Y) ~ stage(X), data= data name)
anova.result <- aov(survival_time~stage,data=clinical_filter)
anova.result
summary_ANOVA<-summary(anova.result)
summary_ANOVA
```
Null hypothesis(H0) of anova test is that there are no differences the survival time among stages.

Alternative hypothesis(HA) is that there are differences in survival time among stages.

p value 0.00661 ** significantly small, thus we can reject the H0 and conclude HA.



Then conduct posthoc analysis to see the difference of a pair of stages.
```{r p1-2}
#TurkeyHSD(aov_result, 'x(predictor variables)', conf.level(optional)= )
posthoc <- TukeyHSD(x=anova.result, 'stage', conf.level=0.95)
print(posthoc)
```
It's safe to say that survival time between (stage 1 and stage 2), (stage 1 and 3) shows meaningful differences based on small enough p-value of results of TukeyHSD function. 

The confidence interval is about [43, 868] days difference for i and ii stages and [152, 1150] days difference for i and iii stages.

### 1-2
```{r p1111}
#t. test(data1, data2, alternative= “two.sided” or “greater” or “less”)
t.test(clinical_filter$survival_time[which(clinical_filter$stage=="stage i")],
       clinical_filter$survival_time[which(clinical_filter$stage=="stage ii")],
       alternative = "greater")
```
#### p-value = 0.00747 is significantly small. stage i patients survival time is greater than stage ii.
#### The confidence interval is about [149.7479 , Inf]

```{r rrr}
t.test(clinical_filter$survival_time[which(clinical_filter$stage=="stage i")],
       clinical_filter$survival_time[which(clinical_filter$stage=="stage iii")],
       alternative = "greater")
```
#### p-value = 0.001406 is sufficiently small. stage i patients survival time is greater than stage iii.
#### The confidence interval is about [296.552 , Inf]

```{r rrrrr}
t.test(clinical_filter$survival_time[which(clinical_filter$stage=="stage ii")],
       clinical_filter$survival_time[which(clinical_filter$stage=="stage iii")],
       alternative = "greater")
```
#### p-value = 0.1151 is not sufficiently small. stage ii patients survival time is not greater than stage iii.
#### The confidence interval is about [ -73.25983 , Inf] which includes zero.

### 1-3
Draw a box plot to see the difference of survival time among the stages.

```{r pressure, echo=FALSE}
boxplot(survival_time~stage, data=clinical_filter)
```

### 1-4~5
```{r rerer}
anova.result <- aov(survival_time~subtype,data=clinical_filter)
anova.result
summary_ANOVA<-summary(anova.result)
summary_ANOVA
```
Null hypothesis(H0) of anova test is that there are no differences the survival time among subtypes.

Alternative hypothesis(HA) is that there are differences in survival time among subtypes.

0.413 is not significantly small, thus we accept H0, posthoc analysis is not needed.
```{r notneed}
posthoc <- TukeyHSD(x=anova.result, 'subtype', conf.level=0.95)

print(posthoc)
```
#### As we expected there's no significantly small p-values. 
#### It's safe to say that different subtypes shows no differences in survival time 
#### based on p-values of results of TukeyHSD function.

### 1-6
Draw a box plot to see the difference of survival time among the subtypes.

```{r pdde, echo=FALSE}
boxplot(survival_time~subtype, data=clinical_filter)
```


## Problem 2
Calculate the number of genes that are differently expressed (DEGs) between group1 and group2.
### answer : 1030
```{r prob2}
#upper quantile and lower quantile patients
x<-quantile(clinical_filter$survival_time, 0.75)
y<-quantile(clinical_filter$survival_time, 0.25)

group1<-dplyr::filter(clinical_filter, survival_time >= x)
group2<-dplyr::filter(clinical_filter, survival_time <= y)

#conduct t test to investigate significant differences
p_val<-c()
for(i in 1:dim(expression_filter)[1]){
  group1_expression<-expression_filter[i,as.character(group1$sample_id)]
  group2_expression<-expression_filter[i,as.character(group2$sample_id)]
  result<-t.test(group1_expression, group2_expression)
  p_val<-c(p_val, result$p.value)}

#Given a set of p-values, returns p-values adjusted using fdr methods
adj_p_val<-p.adjust(p_val, 'fdr')
genes<-rownames(expression_filter)

#diffrently expressed genes (DEGs)
genes_filter1<-genes[which(p_val<0.05)] # chr [1:3835]
genes_filter2<-genes[which(adj_p_val<0.05)]# chr [1:1030] 

#smaller range of genes obtained by adjusted p-values (3835 -> 1030)
```

## Problem 3
Perform survival analysis, to see the difference in survival time depending on stages.


We can conduct survival analysis by using R Function provided by R package “survival”.
```{r rrerererewreq}
library(survival)
```
survdiff(Surv(y, censoring_data)~x)


```{r prob3}
# Kaplan-Meier log-rank test about stage
tt <- as.numeric(clinical_filter$survival_time)/30.5 # survival time in month

delta <- 1*(clinical_filter$vital_status == 0) # censored data
trt <- clinical_filter$stage # stage

survdiff(Surv(tt, delta) ~ trt) # survival analysis depending on stages
```
H0 : There is No difference in survival time among stages

HA : There Exists difference in survival time among stages

With small enough value of p= 0.04, we can conclude HA.


```{r plotttt}
#Visualize the result of Kaplan-Meier log rank test

plot(survfit(Surv(tt, delta) ~ trt), xlab = "Time in months", ylab = "Survival probability",
     col = c("pink", "orange", "violet"), lwd = 2, main = "Survival in Breast Cancer (stage)")

legend("topright", c("stage i", "stage ii", "stage iii"), col = c("pink", "orange", "violet"),
       lty=rep(1,3), lwd=2)
```


Perform survival analysis, to see the difference in survival time depending on subtypes.

```{r prob32222}
# Kaplan-Meier log-rank test about subtypes
tt <- as.numeric(clinical_filter$survival_time)/30.5 # survival time in month

delta <- 1*(clinical_filter$vital_status == 0) # censored data
trt <- clinical_filter$subtype # subtype

survdiff(Surv(tt, delta) ~ trt) # survival analysis depending on stages

```
H0 : There is No difference in survival time among subtypes

HA : There Exists difference in survival time among subtypes

With not enough result p= 0.7 to decide HA, we can conclude H0.


```{r plot44333}
#Visualize the result of Kaplan-Meier log rank test

plot(survfit(Surv(tt, delta) ~ trt), xlab = "Time in months", ylab = "Survival probability",
     col = c("pink", "orange", "violet","red","grey"), lwd = 2, main = "Survival in Breast Cancer (subtype)")
legend("topright", c("Basal-like", "HER2-enriched", "Luminal A", "Luminal B", "Normal-like"), col = c("pink", "orange", "violet","red","grey"),
       lty=rep(1,3), lwd=2)
```

# Problem 4
## Select the gene correlated with survival time
## See the mutation of genes influencing patients' survival time

# 4-1
Analyze the genes in the scope of sifted genes from problem 2.

Calculate pearson correlation coefficient for each genes.


```{r prob4444}
#Correlation test using cor.test function
cor.estimate<-cor.pvalue<-c()
for(i in 1:length(genes_filter2)){
  expression<-expression_filter[genes_filter2[i],]
  survival<-clinical_filter$survival_time[match(clinical_filter$sample_id, colnames(expression_filter))]
  cor.result<-cor.test(expression, survival, method = 'pearson')
  cor.estimate<-c(cor.estimate, cor.result$estimate)
  cor.pvalue<-c(cor.pvalue, cor.result$p.value)}
```
H0 : there is no correlation between 2 variables(DEG, survival time)

=> zero correlation coefficient 

HA : there is correlation between 2 variables(DEG, survival time)

=> non-zero correlation coefficient

*cor.estimate* contains the whole coefficient values calculated from the test.

# 4-2
## Select the data that has p-value less than 0.05

```{r rrrrrrrrrr}
cor.estimate<-cor.estimate[which(cor.pvalue <= 0.05)]
genes<-genes_filter2[which(cor.pvalue <= 0.05)]
```

The most positively correlated gene with survival time, the largest correlation coefficient.

```{r rrrerwerwrew1qqq}
gene<-genes[which(cor.estimate==max(cor.estimate))] # max(cor.estimate) = 0.1377636
gene # "C1orf43"

```
Hence, the gene named "C1orf43" is the one most positively correlated with survival time.

Its Pearson correlation coefficient is 0.1377636.


# 4-3

```{r remaining}

#find the genes that has more than 0.5 Pearson correlation coefficient 
#with the gene "C1orf43"
#Using expression data

#the gene "C1orf43"
```

```{r remaining222}
match("C1orf43", rownames(expression_filter)) # 11928
match("C1orf43", genes_filter2) # 719

for(i in 1:length(genes_filter2)){
  expression<-expression_filter[genes_filter2[i],]
  clorf43<-expression_filter[genes_filter2[719],]
  cor.result<-cor.test(expression, clorf43, method = 'pearson')
  cor.estimate<-c(cor.estimate, cor.result$estimate)
  cor.pvalue<-c(cor.pvalue, cor.result$p.value)}

```
H0 : there is *no* correlation between 2 variables(Gene C1orf43, another gene from gene_filter2)

=> zero correlation coefficient 

HA : there *is* correlation between expression of Gene C1orf43 and another gene from gene_filter2

=> non-zero correlation coefficient

*cor.estimate* contains the whole coefficient values calculated from the test.


## Select the data that has p-value less than 0.05

```{r rrrrrrrr2rr}
cor.estimate<-cor.estimate[which(cor.pvalue <= 0.05)]
genes<-genes_filter2[which(cor.pvalue <= 0.05)]
```

Extract the genes that has more than 0.5 Pearson correlation coefficient with the gene C1orf43,

```{r rrrerwerwr22ew1qqq}
gene<-genes[which(cor.estimate > 0.5)] 
gene #"C1orf43"

```
Hence, the gene "C1orf43" itself is the only gene correlated with expression of itself as more than 0.5.

Its Pearson correlation coefficient is 1.


# Incomplete tasks...

```{r remaininegegeg}
#4-4
#Let’s consider the patients who has mutation for the genes you selected in problem 4- (2),(3) 
#as positive and the others as negative. 
#Perform log rank test between the two groups.






#4-5
#Interpret result from 4-(4) with corresponding p-values 



#Visualize Kaplan-Meier survival curve.


```




```{r four66666}
#4-6
#Which is more efficient to predict survival time 
#among stage, subtype and mutation of genes 
#that was selected in problem 4-(2),(3)?
```
# Above question could be answered by performance analysis